### 2012-08-21 paper_sean/introduction.tex
JMC edits

Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii}, \cite{kohl:stone}. 
It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii}, \cite{clune1}, \cite{clune3}.
However, in most previous works, the experimenter has had to manually decide the regularities of the gaits. 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}. 
However, that work only verified these claims in simulation. 
A follow-up paper evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms \cite{yos:clune}. % insert (figure 1 -- pic of quadratot) 
A further follow-up study \cite{glette} built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware. 
The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs. 180) \cite{yos:clune}, \cite{glette}. 
The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}.

In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of effective regularities \cite{clune2} -- will outperform the manually designed, regularity-constrained, direct encoding once HyperNEAT is combined with a simulator. 
We test this by evolving gaits using HyperNEAT using a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.  
%Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 
%That said, gaits still transferred poorly, revealing that substantial further performance gains are likely via future work that combines a simulator, HyperNEAT, and techniques that reduce the simulator-to-reality gap. 
%More broadly, this work supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. 
%Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

>>>>

Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii}, \cite{kohl:stone}. %JMC - I have a similar claim in my CEC 2009 paper. Grab those cites for here (especially the Hornby ones). 
It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities 
\cite{valsalam:mii,clune2,clune1,clune3}. %JMC-note how you can put multiple cites in one bracket...but no spaces after the comma!
However, in most previous works, the experimenter has had to manually decide the regularities of the gaits. %we need cites for this claim. you can get them from Clune1
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}. 
However, that work only verified these claims in simulation. 
A follow-up paper by Glette et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}. % insert (figure 1 -- pic of quadratot) 
A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware~\cite{glette}. %JMC-Should there be a space between the end of the sentence and the cite? Whatever the answer, we should be consistent. If so, I have added here with the ~. The ~ tells Latex not to split that space from the character before and after it. 
The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and more generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs. 180) \cite{yos:clune}, \cite{glette}. 
The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
%Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}. %JMC- I don't think we need to cite this paper...it is a non-evolutionary method, and it doesn't outperform Glette...and the sentence just sticks out as a detail that isn't necessary to our story.

In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform the manually designed, regularity-constrained, direct encoding once HyperNEAT is combined with a simulator. %JMC-I don't think they used a direct encoding. I prefer describing it as simple and manually designed to produce specific regularities. 
We test this by evolving gaits using HyperNEAT using a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.  %JMC - Try to write without ever saying ``use'' or ``using.'' If you do, you'll find how much it is overused in writing, as evidenced by this last sentence. Often the word is a stand in for a different, clearer verb...and its use is ambiguous. Please search for this word and try to swap it out throughout the rest of the paper. Often times it requires restructuring a sentence, but the sentence usually ends up better. Similarly, make sure to always follow the word ``this'' with referent. This is extremely ambiguous. You know what you refer to when you write the sentence, but readers (often including your future self!) do not know precisely what you mean, and in science precision is critical. 
%Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 
%That said, gaits still transferred poorly, revealing that substantial further performance gains are likely via future work that combines a simulator, HyperNEAT, and techniques that reduce the simulator-to-reality gap. 
%More broadly, this work supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. 
%Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

### 2012-08-21 paper_sean/robot.tex
JMC edits

The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal.
It has 9 degrees-of-freedom (DOF), with 2 DOF for the inner joint and the outer joint of each leg and 1 DOF for the hip joint of the robot. 
The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a ground on which people could compare the results of different gait learning algorithms. 
Currently, there are published data and results from 9 different learning algorithms on the platform.

The body of the robot was printed using an Objet Connex 500 3-D Printing System. 
It was actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
The servos were attached to the robot using off-the-shelf components and parts.
The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree]. 
The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} which would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}. 
The ones most relevant to this study were shutdowns due to high torque and range. 
Often, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command positions which was out of range for the robot \cite{yos:clune} and cause the servos to shutdown.

In order to address this problem and avoid collisions with its own body, a motion cropping system was implemented on the QuadraTot. 
In addition to placing limits on the range of motions for the inner joints, outer joints, and the hip joint by [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] respectively, an extra cropping system named the smart cropping system was adopted from Shen et al. \cite{haocheng} to further prevent servo shutdowns from fierce gaits -- gaits that commanded servo positions to either extremes. 
A leg position is defined by the sum of servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) as represented in Haocheng et al. \cite{haocheng}.
Smart cropping works by modifying the servo positions from fierce gaits, which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
Through experimentation, we found that a \emph{\textbf{g}} of 730 provides a good balance between performance and safety.

The robot was controlled by CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10. 
In previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns. 
The servos were commanded positions using the Pydynamixel library at 40Hz from the computer.

We tracked the motion of the robot using infrared (IR) LEDs and a Wiimote (from the Nintendo Wii gaming system).
Three IR LEDs were placed at the center of the QuadraTot for the Wiimote to detect. 
A Wiimote contains an IR camera which could track positions of IR light. 
For this study, the Wiimote was velcroed onto the ceiling pointing down to maximize its viewable window. 
The Wiimote communicated with the computer tethered to the robot via bluetooth using the CWiid library and a separate Python script recorded the position of the robot reported by the Wiimote during the duration of the run.

>>>>

% !TEX root =main.tex
The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot plans and part files for 3D printing are freely available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal~\cite{yos:clune}.
It has 9 degrees-of-freedom (DOF), with 2 DOF for the inner joint and the outer joint of each leg and 1 DOF for the hip joint of the robot. 
The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a platform with which people could compare the results of different gait learning algorithms. 
Currently, there are published data and results from 9 different learning algorithms on the platform from INSERT different publications. %JMC fill in INSERT, and dump all cites here. This is a good place to cite haochengs paper, which is fun for him and Jason, but does not distract from our story.

The body of the robot was printed using an Objet Connex 500 3-D Printing System. 
It was actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
The servos were attached to the robot using off-the-shelf components and parts.
The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree].

The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} that would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}. 
The ones most relevant to this study were shutdowns due to high torque and range. 
As has been reported previously with QuadraTot\cite{yos:clune}, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command out of range positions and cause the servos to shutdown. In order to address this problem and avoid collisions with its own body, a motion cropping system was implemented on the QuadraTot. 
In addition to placing limits on the range of motions for the inner joints, outer joints, and the hip joint of [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] respectively, an extra cropping system named the smart cropping system was adopted from Shen et al. \cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque. This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC - do they sum the position values, or the joint angles? Since all the previous extremes are expressed in angles, the writing makes me expect angles...
and modifying the servo positions from fierce gaits, which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$. %the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places. No need to invent a definition only to use it 2-3 times...can you swap in different, more descriptive words?
Through experimentation, we found that a \emph{\textbf{g}} of 730 provides a good balance between performance and safety. %JMC - This description is not great. The last sentence's use of g doesn't make sense with the equation...is g a threshold or the sum? Please try to clean up this description.

The robot was controlled by CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10. 
In previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
The servos were commanded positions using the Pydynamixel library at 40Hz from the computer.

We tracked the motion of the robot using infrared (IR) LEDs and a Wiimote (from the Nintendo Wii gaming system).
Three IR LEDs were placed at the center of the QuadraTot for the Wiimote to detect. 
A Wiimote contains an IR camera which could track positions of IR light. 
For this study, the Wiimote was velcroed onto the ceiling pointing down to maximize its viewable window. 
The Wiimote communicated with the computer tethered to the robot via bluetooth using the CWiid library and a separate Python script recorded the position of the robot reported by the Wiimote during the duration of the run\~cite{yos:clune}.

### 2012-08-21 paper_sean/simulator.tex
Updated figure commands, added as many figures as possible.
JMC edits

The simulations were done using a simulator\footnote{Simulator available for Linux and Windows in headless and visual versions at http://quadratot.yosinski.com} designed and written by Kyrre Glette and Jim Torresen in the University of Oslo, Norway\footnote{{kyrrehg, jimtoer}@ifi.uio.no} \cite{glette}. 
It was written in C++ using the NVIDIA PhysX physics software library to provide an accurate physical representation of the behaviors of the robot. 
A physical representation of the QuadraTot was built in simulation to capture the most important parts of the robot like the number and type of joints, mass, and rough shapes and lengths of the robot. 
The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. 
The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.
The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. 
The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. 
In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. 
In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. 
Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=simulator.eps, width=8cm}
\caption[ ]{Physical representation of the QuadraTot in simulation. It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths. The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL. Figure taken from Glette et al. \cite{glette}.}
\end{center}
\end{figure}

>>>>

% !TEX root =main.tex
The simulations were done using a simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} designed and written by Kyrre Glette and Jim Torresen in the University of Oslo, Norway\footnote{{kyrrehg, jimtoer}@ifi.uio.no} \cite{glette}. 
It was written in C++ using the NVIDIA PhysX physics software library to provide an accurate physical representation of the behaviors of the robot. 
A physical representation of the QuadraTot was built in simulation to capture the most important parts of the robot like the number and type of joints, mass, and rough shapes and lengths of the robot. 
The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. 
The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.
The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. 
The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. 
In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. 
In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. 
Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=simulator.eps, width=8cm}
% \caption[ ]{Physical representation of the QuadraTot in simulation. It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths. The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL. Figure taken from Glette et al. \cite{glette}.}
% \end{center}
% \end{figure}
\figp{simulator_outline}{.6}{Physical representation of the QuadraTot in simulation. It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths. The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL. Figure taken from Glette et al. \cite{glette}.}
